A Bidirected String Graph Model for Genome Assembly

### 2012-12-05 presentation/paper.tex
Paper updates

\documentclass[10pt]{article}
%\usepackage{times}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
%\usepackage{threeparttable}
%\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage{listings}

@@

Every living organism contains a collection of hereditary material called a {\em
genome}.  Typically, each individual cell of a multicellular organism contains a
complete copy of the genome.  A genome consists of one or more DNA ({\it
deoxyribonucleic acid}) molecules in the form of {\it chromosomes}.  DNA is a
double-stranded molecule, each strand of which is a polymer of {\em
nucleotides}.  A nucleotide, together with the corresponding nucleotide on the
opposite strand, is called a {\it base pair}.  In DNA, there are four possible
nucleotides: adenine, cytosine, guanine, and thymine, which are abbreviated as
\Base{A}, \Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs
with T on the other strand, and \Base{C} always pairs with \Base{G} on the other
strand.  In addition, each strand has a direction, and the two strands of a DNA
molecule are directed opposite one another.  Furthermore, a DNA molecule may be
circular (as is usually the case for bacterial chromosomes), in which case the
molecule has no ends.

@@

{\it DNA sequencing} is the process of determining the base pair sequence of one
or more DNA molecules.  {\it Genome sequencing} is the use of DNA sequencing to
determine the base pair sequence of an entire genome.

\subsection{Overall workflow for genome sequencing}
\subsection{Genome assemblers}

@@

@@

>>>>

\documentclass[10pt]{article}
%\usepackage{times}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
%\usepackage{threeparttable}
%\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage{hyperref}
%\usepackage{listings}

@@

Every living organism contains a collection of hereditary material called a {\em
genome}.  Typically, each individual cell of a multicellular organism contains a
complete copy of the genome.  A genome consists of one or more DNA ({\it
deoxyribonucleic acid}) molecules in the form of {\it chromosomes}.  DNA is a
double-stranded molecule, each strand of which is a polymer of {\em
nucleotides}.  A nucleotide, together with the corresponding nucleotide on the
opposite strand, is called a {\it base pair} (abbreviation: {\it bp}).  In DNA,
there are four possible nucleotides: adenine, cytosine, guanine, and thymine,
which are abbreviated as \Base{A}, \Base{C}, \Base{G}, and \Base{T},
respectively.  \Base{A} always pairs with T on the other strand, and \Base{C}
always pairs with \Base{G} on the other strand.  In addition, each strand has a
direction, and the two strands of a DNA molecule are directed opposite one
another.  Furthermore, a DNA molecule may be circular (as is usually the case
for bacterial chromosomes), in which case the molecule has no ends.

@@

{\it DNA sequencing} is the process of determining the base pair sequence of
({\it sequencing}) one or more DNA molecules.  {\it Genome sequencing} is the use
of DNA sequencing to determine the base pair sequence of an entire genome.

Sequencing a DNA molecule is very difficult because it exists at the molecular
level, which makes each base pair much too small to be directly observed.
Therefore, indirect laboratory techniques must be used in order to sequence a
DNA molecule.  Advances in these techniques are progressing rapidly and they can
now be performed by automated machines, but virtually every technique suffers
from the limitation that only short fragments of DNA (several dozen to several
thousand bp long) can be sequenced.  Therefore, to sequence an entire genome,
which may be millions of bp long, the DNA from the genome must be broken into
much smaller fragments, then those small fragments must be sequenced.  The
original genome is then computationally reconstructed in a process called {\it
genome assembly}.  {\it Genome sequencing} therefore refers to the entire
process of sequencing a genome, including lab work and computational work, while
{\it genome assembly} only refers to the computational work.

\subsection{Genome assembly}

\label{subsec:reads}

The problem of genome assembly is to reconstruct a genome, as defined earlier,
given a set of substrings from the genome that may originate from any strand of
any DNA molecule in the genome.  Each substring is called a {\it read} and
corresponds to some DNA that was sequenced by the sequencing machine.  In
general, reads may contain errors, including {\it substitutions}, where a base
pair in the read is incorrectly read as some other base pair, as well as {\it
insertions}, where a sequence of one or more random base pairs have been
inserted into some position in the read, and {\it deletions}, where a sequence
of one or more base pairs from the original genome is missing from the interior
of the read.  Reads are expected to be uniformly distributed on the genome with
a certain {\it coverage} {\it coverage} $X$, meaning that each base pair in the
original genome is expected to be included in $X$ different reads, on average.
However, some sequencing technologies produce highly biased data where the
coverage of genome regions is much more variable than that expected by uniform,
random sampling.

Genomes may contain {\it repeats}, which are subsequences that are repeated
multiple times in the genome.  Repeated sequences may be longer than the length
of any individual read, which makes the genome more difficult to assemble.

Reads may contain other information, such as {\it quality scores} or {\it
mate-pair} information, that provide additional information to the assembler.

For the purposes of this paper, the problem of genome assembly is simplified by
assuming error-free reads that are uniformly sampled from the genome.  In
addition, the reads are assumed to be unpaired (meaning that mate-pair
information, if available, is not used).

@@

Genome assembly is superficially similar to the shortest common superstring
(SCS) problem, which is to find the shortest possible string that contains every
string in a given set as substrings.  Genome assembly also requires finding a
longer string that contains all of a set of substrings.  However, the actual
genome may be more than one string, and the genome may be longer than the
shortest possible genome due to repeats.  In addition,  genomes are
double-stranded, and the reads may have the complications described in
\ref{subsec:reads}, such as sequencing error.  Therefore, genome assembly is not
the same as the SCS problem, and in fact seems to be harder, despite the fact
that the SCS problem is already NP-complete\cite{Turner1989}.

In fact, some models of genome assembly have also been proven to be NP-complete
\cite{Medvedev2007}.  However, it is definitely not acceptable to have a genome
assembly algorithm that takes exponential time to complete.  Instead, heuristic
algorithms that tend to produce good results but also tend to have a reasonable
running time must be designed.

@@

This paper explores the genome assembly problem through an algorithm published
in 2005 by Eugene Myers\cite{Myers2005} that is based on the idea of the
fragment string assembly graph.  The algorithm provides a good overview of the
techniques that are used in genome assembly, especially since it draws on ideas
used in previous assemblers, such as the Celera assembler which was used to
assemble the human genome for the first time\cite{Venter2001}.  However, the
algorithm also introduces some new ideas.

The overall algorithm is as follows: first, use overlaps between reads to build
a graph that models the assembly problem.  Next, simplify and analyze this
graph.  Finally, find paths through this graph to reconstruct the original
genome.
